---
title: "covidcountydataR"
output:
  github_document: null
  rmarkdown::html_vignette: null
vignette: >
  %\VignetteIndexEntry{covidcountydataR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Welcome to the R client library for accessing the COVID County Data
database.

Links:

  - [Repository](https://github.com/CovidCountyData/covidcountydataR)
  - [Website](https://covidcountydata.org/)

As of right now, this library is a wrapper around the
[covidcountydata.py](https://github.com/CovidCountyData/covidcountydata.py) Python client. For
more examples and documentation, please see that library. If you are an
R programmer and are willing to contribute to making a native library,
please reach out at our
[repository](https://github.com/CovidCountyData/covidcountydataR)\!

Also, please see the [project website](https://covidcountydata.org) for
more information.

## COVID County Data

Covid County Data (CCD) is a project funded by
[Schmidt Futures](https://schmidtfutures.com/) and seeks to simplify the
data ingestion process for researchers and policy makers who are working
to enact and understand COVID-19 related policies. We accomplish this
goal in several ways:

  - Collect unique, hard-to-acquire, datasets that are not widely
    distributed
  - Aggregate data collected by other related organizations into a
    centralized database
  - Work with other related organizations to expand and improve their
    data collection processes
  - Build tools, such as this library (and
    [Python](https://github.com/CovidCountyData/covidcountydata.py) and
    [Julia](https://github.com/CovidCountyData/CovidCountyData.jl) equivalents), to
    simplify the data ingestion process

More information about our project and what data is collected can be
found on our [website](https://covidcountydata.org/).

We are always looking to hear from both those who would like to help us
build CCD and those who would like use CCD. [Please reach out to
us](https://covidcountydata.org/contact)\!

## Installation

Please install this package using `devtools::install_github` as follows

```{r install, eval=FALSE}
devtools::install_github("CovidCountyData/covidcountydataR")
```

After installing the package, you need to make sure that the underlying
python package is installed.

To do this, use

```{r installPy, eval=FALSE}
covidcountydataR::install_ccdPY()
```

During the installation process, R will check if you have an existing
Python installation that can be used

You may be prompted to accept the installation of a dedicated Python
(via miniconda) for R to use

We recommend that you accept this request, but if you are comfortable
managing your own Python installation you can say no

## Creating a Client

Once the package is installed, the first step is to create an API
client:

```{r setup}
library(covidcountydataR)
```

```{r createClient}
cl <- client()
```

## Datasets

You can see a list of currently available datasets using:

```{r datasets}
datasets(cl)
```

Each dataset has an associated function

You can get detailed information on a specific dataset using the `info`
method. For example

```{r info}
info(cl)

info(cl, "demographics")

info(cl, "covid_historical")
```

## Requesting Data

Requesting a dataset has three parts:

1.  Create a client
2.  Build a request with desired datasets
3.  `fetch` the datasets

### 1\. Create a client

To create a client, use the `client` function as shown above

```{r client2, eval=FALSE}
cl <- client()
```

You can optionally pass in an API key if you have one (see section on
API keys below)

```{r clientWithKey, eval=FALSE}
cl <- client("my api key")
```

If you have previously registered for an API key (again, see below) on
your current machine, it will be loaded and used automatically for you

In practice you should rarely need to pass the apikey by hand unless you
are loading the key from an environment variable or another source

### 2. Build a request

Each of the datasets in the API have an associated function

To add datasets to the current request, `datasetName(client)` function:

```{r buildRequestFuncs}
covid_us(cl, state="CA")

demographics(cl)

cl
```

You can see that the printed form of the client is updated to show you
what the current request looks like

To clear the current request, use `reset(cl)`:

```{r resetClient}
reset(cl)
#> CCD Client
```

Each dataset function will build up a request for the client and will
return the client itself

This allows us to use the pipe operator (`%>%`) to do the above as:

```{r buildDatasetPipes}
cl %>% covid_us(state="CA") %>% demographics()
```

#### Filtering data

Each of the dataset functions has a number of filters that can be
applied

This allows you to select certain rows and/or columns

For example, in the above example we had `covid_us(state="CA")`. This
instructs the client to only fetch data for counties in the state of
California

Refer to the `info` for each dataset’s function for more information on
which filters can be passed

Also, check out the examples section at the end for more examples

**NOTE:** If a filter is passed to one dataset in the request but is
applicable to other datasets in the request, it will be applied to *all*
datasets

For example in `cl %>% covid_us(state="CA") %>% demographics()` we only
specify a `state` filter on the `covid_us` dataset

However, when the data is collected it will also be applied to
`demographics`

We do this because we end up doing an inner join on all requested
datasets, so when we filter the state in `covid_us` they also get filtered
in `demographics`

### 3\. Fetch the data

Now for the easy part\!

When you are ready with your current

To fetch the data, call the `fetch` function on the client:

```{r fetch1}
df <- fetch(cl)
df

names(df)
```

Notice that after each successful request, the client is reset so there
are no “built-up” requests:

```{r clearedClientAfterFetch}
cl
```

## API keys

Our API is and always will be free for unlimited public use

However, we have an API key system in place to help us understand the
needs of our users

We kindly request that you register for an API key so we can understand
how to prioritize future work

In order to do so, you can use the `register` function

```{r register, eval=FALSE}
register(cl)
```

By default, function will prompt you to input an email address

You can also pass the email address as the second argument for
non-interactive use

```{r registerwithEmail, eval=FALSE}
register(cl, "me@me.com")
```

After you `register` for an API key it will be added to the client. All
future requests with this client will use the API key

We also save the key to a file at `~/.covidcountydata/apikey`

If this file exists, each time you call `client` and do not explicitly
pass an apikey we will read the key from `~/.covidcountydata/apikey` and
automatically apply it for you

Thus, to use the key in future sessions you just need to do `cl <-
client()` and we’ll handle the key for you\!

## Final thoughts

Due to the urgency of the COVID-19 crisis and the need for researchers,
modelers, and policy makers to have accurate data quickly, this project
moves fast\!

We have created this library so that as we add new datasets to our
backend, they automatically appear here and are accessible via this
library

Please check back often and see what has been updated

### Examples

```{r singleDatasetAll, cache=TRUE}
# Single dataset all
cl %>% mobility_devices() %>% fetch()
```

```{r singleDatasetDeathFilter}
# Single dataset filter on deaths
cl %>% covid_us(location="<100", variable="deaths_total", value=">100") %>% fetch()
```

```{r stateFilter}
# Single dataset single states with all counties
# OR: `cl %>% mobility_devices(state=as.integer(48)) %>% fetch()`
# OR: `cl %>% mobility_devices(state="TX") %>% fetch()`
cl %>% mobility_devices(state="48") %>% fetch()
```

```{r multipleStates}
# Single dataset multiple states with all counties
cl %>% mobility_devices(state=c("CA", "TX")) %>% fetch()
```

```{r variableSelection}
# Single dataset variable select
cl %>% demographics(variable = c("Total population", "Fraction of population over 65", "Median age")) %>% fetch()
```

```{r multpleDatasets, cache=TRUE}
# Multiple datasets all data
cl %>% demographics() %>% covid_us(dt=">2020-07-20") %>% fetch()
```

```{r multipleOnlyStates, cache=TRUE}
# Multiple datasets states only
cl %>% demographics() %>% covid_us(location="<100", dt=">2020-07-20") %>% fetch()
```

```{r multipleOnlyCounties, cache=TRUE}
# Multiple datasets counties only
cl %>% demographics() %>% covid_us(location=">1000", dt=">2020-07-20") %>% fetch()
```
